---
title: "homework_3"
author: "Hambee Makinoda"
date: 2026-03-04
format:
  html:
    toc: true 
execute:
  message: false 
  warning: false 
---

**GitHub Repository:** [View Repository](https://github.com/Hambanana0923/ENVS-193DS_homework-03)

# Set Up

```{r}
#| label: reading in packages and data
#| message: false

library(tidyverse)
library(here)
library(janitor)
library(readxl)

salinity <- read_csv("data/salinity-pickleweed.csv")
personal <- read_csv("data/env193ds_personaldata.csv")
```

# Problem 1. Slough soil salinity

## a. An appropriate test

Two appropriate tests are Pearson's correlation and Spearman's rank correlation, which measures the STRENGTH of the relationship between soil salinity and California pickleweed biomass. Pearson's correlation uses two continuous variables that were observed independently, and assumes that the variables are normally distributed to see how strongly two variables change together in a straight-line/linear relationship. Spearman's rank correlation looks for monotonic relationship between variables in independent observations without assuming normality, meaning they evaluate whether or not biomass generally increases or decreases with salinity.

## b. Create a visualization

```{r}
#| label: Visualization

# base layer: ggplot call
ggplot(data = salinity, # use salinity data frame
       aes(x = salinity_mS_cm, # set x axis to salinity_mS_cm column 
           y = pickleweed)) + # set y axis to pickleweed column
  geom_point(color = "darkgreen") + # first layer: scatter plot & change color to dark green
  labs(
    x = "Soil Salinity (mS/cm)", # change x-axis label
    y = "California Pickleweed Biomass (g)", # change y-axis label
    title = "Relationship Between Soil Salinity and Pickleweed Biomass") + # name the title of the graph
  theme_minimal() # change ggplot theme to minimal
```

## c. Check your assumptions and run your test

Checking assumption:

```{r}
#| label: salinity obs (qqplot normality check)

ggplot(data = salinity, # starting data frame
       aes(sample = salinity_mS_cm)) + # y-axis for a QQ plot
  geom_qq_line(color = "blue") + # reference line in blue
  geom_qq() + # actual points for the QQ plot
  theme_minimal() + # changing the theme to minimal
  theme(legend.position = "none") #taking out legend

```

```{r}
#| label: biomass obs (qqplot normality check)

ggplot(data = salinity, # starting data frame
       aes(sample = pickleweed)) + # y-axis for a QQ plot
  geom_qq_line(color = "red") + # reference line in blue
  geom_qq() + # actual points for the QQ plot
  theme_minimal() + # changing the theme to minimal
  theme(legend.position = "none") #taking out legends
```

I checked the assumptions of normality and linearity for Pearson’s correlation. I used QQ plots to visually assess whether the distributions of soil salinity (mS/cm) and pickleweed biomass (g) were approximately normal, and I used a scatterplot from part b to visually check whether the relationship between the variables appeared linear. The QQ plots for both salinity and biomass showed individual observations roughly following the reference line with very little outliers and the scatterplot suggested a roughly linear relationship, indicating that this data met all appropriate assumptions.

------------------------------------------------------------------------

Running Pearson's correlation test

```{r}
#| label: Pearson's correlation

cor.test( # correlation test call
  salinity$salinity_mS_cm, # choose salinity column for one variable
  salinity$pickleweed, # choose pickleweed column for another variable
  method = "pearson") # choose parametric correlation


```
