---
title: "homework_3"
author: "Hambee Makinoda"
date: 2026-03-04
format:
  html:
    toc: true 
execute:
  message: false 
  warning: false 
---

**GitHub Repository:** [View Repository](https://github.com/Hambanana0923/ENVS-193DS_homework-03)

# Set Up

```{r}
#| label: reading in packages and data
#| message: false

library(tidyverse)
library(here)
library(janitor)
library(readxl)

salinity <- read_csv("data/salinity-pickleweed.csv")
personal <- read_csv("data/env193ds_personaldata.csv")
```

# Problem 1. Slough soil salinity

## a. An appropriate test

Two appropriate tests are Pearson's correlation and Spearman's rank correlation, which measures the STRENGTH of the relationship between soil salinity and California pickleweed biomass. Pearson's correlation uses two continuous variables that were observed independently, and assumes that the variables are normally distributed to see how strongly two variables change together in a straight-line/linear relationship. Spearman's rank correlation looks for monotonic relationship between variables by ranking the values for each variables separately in independent observations without assuming normality, meaning they evaluate whether or not biomass generally increases or decreases with salinity.

## b. Create a visualization

```{r}
#| label: Visualization

# base layer: ggplot call
ggplot(data = salinity, # use salinity data frame
       aes(x = salinity_mS_cm, # set x axis to salinity_mS_cm column 
           y = pickleweed)) + # set y axis to pickleweed column
  geom_point(color = "darkgreen") + # first layer: scatter plot & change color to dark green
  labs(
    x = "Soil Salinity (mS/cm)", # change x-axis label
    y = "California Pickleweed Biomass (g)", # change y-axis label
    title = "Relationship Between Soil Salinity and Pickleweed Biomass") + # name the title of the graph
  theme_minimal() # change ggplot theme to minimal
```

## c. Check your assumptions and run your test

Checking assumption:

```{r}
#| label: salinity obs (qqplot normality check)

ggplot(data = salinity, # starting data frame
       aes(sample = salinity_mS_cm)) + # y-axis for a QQ plot
  geom_qq_line(color = "blue") + # reference line in blue
  geom_qq() + # actual points for the QQ plot
  theme_minimal() + # changing the theme to minimal
  theme(legend.position = "none") #taking out legend

```

```{r}
#| label: biomass obs (qqplot normality check)

ggplot(data = salinity, # starting data frame
       aes(sample = pickleweed)) + # y-axis for a QQ plot
  geom_qq_line(color = "red") + # reference line in blue
  geom_qq() + # actual points for the QQ plot
  theme_minimal() + # changing the theme to minimal
  theme(legend.position = "none") #taking out legends
```

I checked the assumptions of normality and linearity for Pearson’s correlation. I used QQ plots to visually assess whether the distributions of soil salinity (mS/cm) and pickleweed biomass (g) were approximately normal, and I used a scatterplot from part b to visually check whether the relationship between the variables appeared linear. The QQ plots for both salinity and biomass showed individual observations roughly following the reference line with very little outliers and the scatterplot suggested a roughly linear relationship, indicating that this data met all appropriate assumptions.

------------------------------------------------------------------------

Running Pearson's correlation test

```{r}
#| label: Pearson's correlation

cor.test( # correlation test call
  salinity$salinity_mS_cm, # choose salinity column for one variable
  salinity$pickleweed, # choose pickleweed column for another variable
  method = "pearson") # choose Pearson's correlation


```

## d. Results communication

[Which test used and why:]{.underline}\

To evaluate the strength of the relationship between California pickleweed biomass (g) and soil salinity (mS/cm), I used a Pearson’s correlation test because both variables are continuous & normally distributed, and I wanted to measure the strength of their relationship.

[Interpretation of the test:]{.underline}\

The results show a moderate positive relationship between soil salinity and pickleweed biomass, meaning that pickleweed biomass tends to increase as soil salinity increases. The correlation coefficient was moderate and positive (Pearson's r = 0.53, t(21)= 2.90, p = 0.0086, $\alpha$ = 0.05), suggesting that higher salinity conditions are associated with greater pickleweed biomass in this restoration site.

## e. Test implications

Our analysis shows a moderate and positive relationship between soil salinity and pickleweed biomass, meaning pickleweed tends to grow larger in areas with higher salinity levels. This statistical result suggests that restoration efforts of pickleweed may be more successful in parts of the site that are more saline, since pickleweed appears to have larger biomass at more saline conditions. When planning new plantings, we should prioritize areas with similar, high salinity levels to improve the chances of strong plant growth.

## f. Double check your own work

```{r}
#| label: Spearman's correlation

cor.test( # correlation test call
  salinity$salinity_mS_cm, # use salinity data
  salinity$pickleweed, # use pickleweed biomass data
  method = "spearman") # choose Spearman's correlation

```

The Spearman’s correlation also shows a significant positive relationship between soil salinity (mS/cm) and pickleweed biomass (g) (Spearman ρ = 0.593, S = 824, p = 0.0034, $\alpha$ = 0.05). This leads to the same conclusion as the Pearson correlation test (Pearson's r = 0.53, t(21)= 2.90, p = 0.0086, $\alpha$ = 0.05): pickleweed biomass tends to increase as soil salinity increases. While Pearson evaluates a linear relationship using the raw values, Spearman evaluates a monotonic relationship using ranked values, but both tests indicate a similar positive association between the variables.

# Problem 2. Personal data

## a. Updating your visualizations

```{r}
#| label: categorical

personal_clean <- personal |> # start with storing personal data frame in personal_clean
  clean_names() # clean names so it only contains lowercase and underscore

ggplot(data = personal_clean, # start ggplot call with my_data_clean data frame
       aes(x = day_type, # choose x-axis
           y = screen_time_min,# choose y-axis
           color = day_type)) + # change color based on day_type 
  geom_boxplot() + # first layer: boxplot
  geom_jitter(height = 0, # second layer: jitter
              width = 0.2, # no height jitter, limit horizontal jitter
              alpha = 0.7) + # lower opacity of jitter
  labs(title = "Screen Time Differs between School Day and Non-School Day", # label title
       subtitle = "2026-03-03", # label subtitle with most recent observation date
       x = "Day Type", # relabel x-axis
       y = "Screen Time (min)") + # relabel y-axis
  theme_classic() + # change ggplot theme
  theme( # use theme() function
    plot.title = element_text (size = 11), # change title size
    plot.subtitle = element_text (size = 10)) + # change subtitle size 
    scale_color_manual(values = c( # Manually select colors of the graph 
      "School" = "darkgreen",
      "Non-School" = "brown4" )) # manually change colors of the graph to be different from default
```

------------------------------------------------------------------------

```{r}
#| label: continuous 

ggplot(data = personal_clean, # start with my_data_clean data frame
       aes(x = amount_of_sleep_min, # choose continuous variable x-axis
           y = screen_time_min)) + # choose response variable for y-axis
  geom_point(color = "yellow3") + # first layer: scatterplot & change point color to red 
  labs(title = "Relationship between Sleep Time and Screen Time", # label title 
       subtitle = "2026-03-03", # label subtitle with most recent observation date
       x = "Amount of Sleep (min)", # relabel x-axis
       y = "Screen Time (min)") + # relabel y-axis
  theme_classic() # change ggplot theme 
```
